---
title: "Final Project"
output:
  html_document:
    df_print: paged
---

### Does Nature Know Best? Analyzing Activity in Nature & Mental Health

This final project investigates the relationship between activity in nature and mental health. First a chi-squared test is conducted, followed by three logistic regressions with one of the models accounting for two confounding variables.

**Hypothesis**: Individuals who are active in nature are more likely to report better mental health.

This project is based on the General Social Survey (GSS) 2021 data is found here: <https://gss.norc.org/Get-The-Data>

```{r}
#import libraries and more libraries
library(tidyverse)
library(dplyr)
library(descr)
library(haven)
library(DescTools)
library(vcd)
library(ggplot2)
library(infer)
library(wesanderson)
library(lmtest)
```

#### Load GSS Data

```{r}
gss_2021 <- read_dta("gss2021.dta")

#view(gss_2021)
#colnames(gss_2021)
```

#### Select Variables and Drop NAs

```{r}
#select variables for analysis and drop NAs
data <- gss_2021 %>% 
  select(id, hlthmntl,incom16, activnat, hlthphys, income16) %>% drop_na(hlthmntl, activnat)
data 
```

#### Variable Histograms and Counts

Visualize the data and see counts for categories

```{r}
#histograms
data %>%  ggplot(aes(x=activnat )) + geom_histogram()

#counts of nature variables
data %>% count(activnat)
```

## Chi-Squared Test

The association between the categorical variables activnat and hlthmntl was highly significant with a p-value of 0.000013080. A similar story is reflected in the simulation-based null distribution visualization.

```{r}
#test data type
sapply(data, typeof)

#recode columns
data$activnat <- as.character(data$activnat)
data$hlthmntl <- as.character(data$hlthmntl)

#active in nature
observed <- data %>% 
    specify(hlthmntl ~ activnat) %>% 
    hypothesize(null = "independence") %>% 
    calculate(stat = "Chisq") #chi squared distribution
observed

data %>% 
    specify(hlthmntl ~ activnat) %>% 
    hypothesize(null = "independence") %>% 
    generate(reps = 1000, type = "permute") %>%
    calculate(stat = "Chisq") %>% 
    visualize() +
    shade_p_value(observed, direction = "greater") 

chisq_test(x = data, hlthmntl ~ activnat)
```

#### Mosaic Plot and Pearson Residuals

Using a mosaic plot to visualize the cross between both variables and spot the Pearson Residuals.

```{r}
#create a mosaic plot from the table
tbl <- xtabs(~hlthmntl + activnat, data)
ftable(tbl)

mosaic(tbl, main = "Active in Nature Impact on Mental Health") 

#more detailed mosaic
mosaic(tbl, 
       shade = TRUE,
       legend = TRUE,
       labeling_args = list(set_varnames = c(activnat = "Activity in Nature",
                                             hlthmntl = "Mental Health")),
       main = "Activity in Nature Impact on Mental Health")
```

#### Convert Mental Health into Binary (Two Ways)

Applying to thought processing to split the hlthmntl variable (mental health) into a binary.

```{r}
#recoding that only poor is compared to grouped other
data_poor <- data %>% 
  mutate(mntl_poor = case_when(
  hlthmntl == 1 ~ 1,
  hlthmntl == 2 ~ 1,
  hlthmntl == 3 ~ 1,
  hlthmntl == 4 ~ 1,
  hlthmntl == 5 ~ 0,
))
data_poor

#recoding the poor and fair grouped vs. all other
data_recode <- data %>% 
  mutate(mntl_binary = case_when(
  hlthmntl == 1 ~ 1,
  hlthmntl == 2 ~ 1,
  hlthmntl == 3 ~ 1,
  hlthmntl == 4 ~ 0,
  hlthmntl == 5 ~ 0,
))
data_recode

#visualization code for below (listed as a character)
data_text <- data %>% 
  mutate(mntl_binary1 = case_when(
  hlthmntl == 1 ~ "good",
  hlthmntl == 2 ~ "good",
  hlthmntl == 3 ~ "good",
  hlthmntl == 4 ~ "bad",
  hlthmntl == 5 ~ "bad",
  TRUE~ as.character((hlthmntl)
)))
data_text

#convert back to numeric
data_recode$hlthmntl <- as.numeric(data_recode$hlthmntl)

#mental health original
data_recode %>%  ggplot(aes(x= hlthmntl)) + geom_bar(fill = "#3D550C" )

#mental health binary split
data_poor %>%  ggplot(aes(x= mntl_poor)) + geom_bar(fill = "#81B622")

#mental health split - poor only
data_recode %>%  ggplot(aes(x= mntl_binary)) + geom_bar(fill = "#81B622")
```

#### Bar Chart Visualization for activnat and mental_binary1

Examine the two variables as a proportional bar chart and bar chart.

```{r}
#proportional bar chart
data_text %>% 
  ggplot(aes(x = activnat, fill = mntl_binary1)) +
  geom_bar(position="fill") + scale_fill_manual(values = wes_palette(2, name = "GrandBudapest1", type = "continuous"))

#bar chart
data_text %>% 
  ggplot(aes(x = activnat, fill = mntl_binary1)) +
  geom_bar(notch = TRUE) + scale_fill_manual(values = wes_palette(2, name = "GrandBudapest1", type = "continuous"))
```

## Logistic Regression \#1

```{r}
summary(glm(mntl_binary~factor(activnat), family=binomial, data = data_recode)) #active in nature and mental health binary

#summary(glm(mntl_poor~factor(activnat), family=binomial, data = data_poor))#active in nature vs. poor health (NOT INCLUDED IN SLIDES)
```

#### Probabilities for Logistic Regression \#1: Active in Nature and Mental Health Binary

```{r}
#daily (cat. 1)
exp(2.1180)/(exp(2.1180)+1)

#several times a week (cat. 2)
exp(2.1180 + -0.2604)/(exp(2.1180+ -0.2604)+1)

#several times a month (cat. 3)
exp(2.1180 + -0.4465)/(exp(2.1180+ -0.4465)+1)

#several times a year (cat. 4)
exp(2.1180 + -0.6996)/(exp(2.1180+ -0.6996)+1)

#never (cat. 5)
exp(2.1180 + -0.8728)/(exp(2.1180+ -0.8728)+1)

0.8926404/0.7764679-1 #cat. 1 is bigger than cat.5 by this much
```

## Logistic Regression \#2: Making Independent Variables for Activity in Nature Continuous

1\. DAILY 1 -\> 365 days

2\. SEVERAL TIMES A WEEK 2 -\> 4\*4\*12 = 192 days

3\. SEVERAL TIMES A MONTH 3 -\> 4\*12 = 48 days

4\. SEVERAL TIMES A YEAR 4 -\> 4\*1 = 4 days

5\. NEVER 5 -\> 0 days

```{r}
#convert to continuous 
data_nature <- data_recode %>% 
  mutate(nature_cont = case_when(
  activnat == 1 ~ 365,
  activnat == 2 ~ 192,
  activnat == 3 ~ 48,
  activnat == 4 ~ 4,
  activnat == 5 ~ 0,
))
data_nature

#test out logistic regressions
summary(glm(mntl_binary~factor(activnat), family=binomial, data = data_recode)) #active in nature and mental health binary

summary(glm(mntl_binary~nature_cont, family=binomial, data = data_nature)) #active in nature and mental health binary
```

#### Further Testing Data and Probabilities

```{r}
#y= beta zero + beta one (days in nature)
-1.4743232 + -0.0018971*48
-1.4743232 + -0.0018971*192

0.2289336/(0.2289336+1) #intercept
0.2284997/(0.2284997+1)
0.03641082/( 0.03641082+1)

exp(-1.4743232)
exp(-0.0018971)*exp(-1.4743232)
exp(-1.838566)*exp(-1.4743232)

data_recode %>% count(mntl_binary)

table(data_recode$mntl_binary, data_recode$activnat) #create table
```

#### Calculating Probability for Days in Nature

```{r}
#48 days
0.0918602+(0.0018971*48)
exp(0.182921)/(exp(0.182921)+1) #probability

#192 days
0.0918602+(0.0018971*192)
exp(0.4561034)/(exp(0.4561034)+1) #probability

#365 days
0.0918602+(0.0018971*365)
exp(0.7843017)/(exp(0.7843017)+1) #probability
```

#### Making Confounding Variable (Income16) Continuous

```{r}
#convert to numeric
data_nature$income16 <- as.numeric(data_nature$income16)

data_coded <- data_nature %>% drop_na(income16) %>% 
  mutate(income = case_when(income16 == "1" ~ 500,
                            income16 == "2" ~ 1999.5,
                            income16 == "3" ~ 3499.5,
                            income16 == "4" ~ 4499.5,
                            income16 == "5" ~ 5499.5,
                            income16 == "6" ~ 6499.5,
                            income16 == "7" ~ 7499.5,
                            income16 == "8" ~ 8999.5,
                            income16 == "9" ~ 11249.5,
                            income16 == "10" ~ 13749.5,
                            income16 == "11" ~ 16249.5,
                            income16 == "12" ~ 18749.5,
                            income16 == "13" ~ 21249.5,
                            income16 == "14" ~ 23749.5,
                            income16 == "15" ~ 27499.5,
                            income16 == "16" ~ 32499.5,
                            income16 == "17" ~ 37499.5,
                            income16 == "18" ~ 44999.5,
                            income16 == "19" ~ 54999.5,
                            income16 == "20" ~ 67499.5,
                            income16 == "21" ~ 82499.5,
                            income16 == "22" ~ 99999.5,
                            income16 == "23" ~ 119999.5,
                            income16 == "24" ~ 139999.5,
                            income16 == "25" ~ 159999.5,
                            income16 == "26" ~ 200000)) #assumed for highest amount since it's $170,000 OR OVER
data_coded
```

#### Making Confounding Variable (Physical Health) Binary

```{r}
#recoding the poor and fair grouped vs. all other
data_confounder <- data_coded %>% 
  mutate(phys_binary = case_when(
  hlthphys == 1 ~ 1,
  hlthphys == 2 ~ 1,
  hlthphys == 3 ~ 1,
  hlthphys == 4 ~ 0,
  hlthphys == 5 ~ 0
))
data_confounder
```

## Logistic Regression \#3: Adding Confounding Variables

```{r}
#factor categorical version of activnat
summary(glm(mntl_binary~factor(activnat) + factor(phys_binary) + income, family=binomial,data = data_confounder)) 

#continuous version of activnat
summary(glm(mntl_binary ~ nature_cont + factor(phys_binary) + income, family=binomial,data = data_confounder))  #NOT IN FINAL SLIDES
```

#### Logistic Regression \#3 Probabilities

```{r}
exp(0.042762183+(0.000008319*4499.5))/(exp( 0.042762183+(0.000008319*4499.5))+1) #mean value of originally category 4
exp(0.042762183+(0.000008319*139999.5))/(exp( 0.042762183+(0.000008319*139999.5))+1) #mean value of originally category 24

exp(0.042762183+(0.000008319*40000))/(exp( 0.042762183+(0.000008319*40000))+1) #$40K income
exp(0.042762183+(0.000008319*50000))/(exp( 0.042762183+(0.000008319*50000))+1) #$50K income
exp(0.042762183+(0.000008319*60000))/(exp( 0.042762183+(0.000008319*60000))+1) #$60k income
exp(0.042762183+(0.000008319*70000))/(exp( 0.042762183+(0.000008319*70000))+1) #$70k income
exp(0.042762183+(0.000008319*80000))/(exp( 0.042762183+(0.000008319*80000))+1) #$80k income
exp(0.042762183+(0.000008319*100000))/(exp( 0.042762183+(0.000008319*100000))+1) #$100k income

```

#### Logistic Regression: Confounding Variables with Interaction Coefficients

These models were not included in the final presentation because none of the interaction coefficients were insignificant.

```{r}
#factor categorical version of activnat
summary(glm(mntl_binary~factor(activnat) * factor(phys_binary) + factor(activnat) * income, family=binomial,data = data_confounder)) #NOT IN FINAL SLIDES

#continuous version of activnat
summary(glm(mntl_binary ~ nature_cont * factor(phys_binary) + nature_cont * income, family=binomial,data = data_confounder))
```

## Testing Pseudo R-Squared for Logistic Regression \#1 vs. \#3

```{r}
#testing Pseudo R-Squared
r.glm  <- glm(mntl_binary~factor(activnat),family=binomial, data = data_recode)
PseudoR2(r.glm, which = "McFadden")

r.glm  <- glm(mntl_binary~factor(activnat)*factor(phys_binary) + activnat*income,family=binomial, data = data_confounder)
PseudoR2(r.glm, which = "McFadden")
```

## Likelihood Ratio Test: Logistic Regression \#1 vs. \#3

```{r}
options(scipen = 900) #get rid of scientific notation

#drop any additional NAs
data_model <- data_confounder %>% drop_na(activnat,phys_binary,mntl_binary, income, incom16)

#fit full model
model_full <- glm(mntl_binary ~ factor(activnat) + phys_binary + income, data = data_model)

#fit reduced model
model_reduced <- glm(mntl_binary ~ factor(activnat), data = data_model)

#perform likelihood ratio test for differences in models
lrtest(model_full, model_reduced)
```
