---
title: "Homework 07 - Schreiber"
output:
  html_document:
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# Homework 7

### Import Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#import libraries
library(tidyverse)
library(ggplot2)
```

### Load Data

```{r}
titanic <- read_csv("titanic3.csv") #import data

titanic1 <- titanic %>% #make headings lower case
  rename_with(tolower) 
```

```{r}
mean(titanic1$survived) #38% of people survived in this sample.
```

## Question 1

*1. Construct a table showing the distribution of passengers by class
and survival.*

Here's the table that shows the distribution of passengers by class and
whether the people survived or not. The column numbers (1 through 3)
represent the three classes on the Titanic and the row numbers refer to
survived (1) and not survived (0).

```{r}
table(titanic1$survived, titanic1$pclass)
```

## Question 2

*2. Construct a logistic regression model that links survival to the
passenger class. Write out the equation first without Running it in R.
HINT: Class is a factor variable*

Assuming the default is first class:

$$ logODDs (survived) = β0 + (β1 * class2) + (β2 * class3)
$$

## Question 3

*3. Using hand-calculations, determine the coefficients in the model and
interpret them (HINT: all you need to do is to use the table, calculate
odds for the default category and the odds-ratios for the other
categories versus the default)*

$$
logODDs (survived) = β0 + (β1 * class2) + (β2 * class3)
$$

$$
logODDs (survived) = 0.486133 + (- 0.7696045 *class2) + (- 1.556732*class3)
$$

The odds for first class (default category) was 1.62, while the odds
ratio for second class was 0.46 and for third class was 0.21. The number
0.486133 in the model represents the log-odds survival of the default
category first class. The number -0.7696045 in the model is calculated
by the difference between the log-odds of second class minus the
log-odds of first class. Lastly, the number -1.556732 in the above model
is calculated by the difference between the log-odds of third class
minus the log-odds of the first class.

Of note, the odds of survival in people in class 2 is 53.7% lower than
class 1 and the odds of survival in people in class 3 is 78.9% lower
than class 1.

### Odds and Odds Ratios for Class

```{r}
#Odds of Survival for First Class (default category)
200/123

#Odds Ratio for Second Class
(119/158)/(200/123)
((119/158)/(200/123)-1)*100 #odds of survival in people in class 2 is 53.68% lower than class 1

#Odds Ratio for Third Class
(181/528)/(200/123)
((181/528)/(200/123)-1)*100  #odds of survival in people in class 3 is 78.91% lower than class 1
```

### Log Odds for Class

```{r}
#Log Odds of Survival for First Class
log(200/123)

#Log Odds of Survival for Second Class
log(119/158) #note that it's negative

#Log Odds of Survival for Third Class
log(181/528)

#Difference to calculate betas
-0.2834715 - 0.486133 #beta one (Log Odds of Second Class - Log Odds of First Class)

-1.070599 - 0.486133 #beta two (Log Odds of Third Class - Log Odds of First Class)
```

## Question 4

*4. Now Run the model in R. Confirm that you got the same results as in
part c). Interpret the results and talk about significance (both
statistical and substantive).*

$$
logODDs (survived) = 0.4861 + (-0.7696 *class2) + (-1.5567 *class3)
$$

Taking into account the slight number variances due to rounding in the
model, the numbers here are similar to the answer provided in question
3. There is a difference in survival for each variable and that
difference is statistically significant for each class (each p-value is
below the 0.05 threshold).

You're less likely to survive in class 2 and class 3 than class 1.
You're also less likely to survive in class 3 than class 2. First class
(the default category) has odds of 1.62 for survival on the Titanic.
Additionally, the odds of survival for class 2 was 0.75 and 0.34 for
class 3. This means that class 1 had the highest odds of survival and
class 3 ended up with the lowest odds of survival.

Additionally, the odds ratio for second class is about 0.46, which means
that there is 53.7% lower odds of survival compared to first class. For
third class, the odds ratio is 0.21, which means that there is a 78.9%
lower odds of survival compared to first class. Both of these lower odds
for class 2 and class 3 compared to class 1 are substantively lower.

### Logistic Regression

```{r}
summary(glm(survived~factor(pclass), family = "binomial", data = titanic1))
```

### Odds Calculations

```{r}
#convert to odds
exp(0.4861) #intercept class 1

#convert to odds ratio
exp(-0.7696)
exp(-1.5567)

exp(-0.7696)*exp(0.4861) #class 2 odds of survival
exp(-1.5567)*exp(0.4861) #class 3 odds of survival

(0.4631983-1)*100 #odds of survival in people in class 2 is 53.68% lower than class 1
(0.21083070-1)*100 #odds of survival in people in class 3 is 78.91% lower than class 1
```

## Question 5

*5. What's the probability of survival for each class of passengers?*

Probability of *First Class Survival*: 0.619195 or about 61.9%

Probability of *Second Class Survival*: 0.4296029 or about 43.0%

Probability of *Third Class Survival*: 0.2552891 or about 25.5%

### Probability of Survival for Each Class of Passengers

```{r}
#Probability of First Class Survival
200/(200+123) #first way
exp(0.4861)/(exp(0.4861)+1) #second way

#Probability of Second Class Survival
119/(158+119) #first way
exp(0.4861 + -0.7696)/(exp(0.4861 + -0.7696)+1) #second way

#Probability of Third Class Survival
181/(528+181) #first way
exp(0.4861 + -1.5567)/(exp(0.4861 + -1.5567)+1) #second way
```

## Question 6

*6. Construct a model that interacts class of passenger and his/her
gender. Interpret the results the same way you did before.*

**Equation and Model**

$logODDs (survived)= β0 + (β1 * class2) + (β2 * class3) + (β3 * male) + (β4 * male * class2) + (β5 * male * class3)$

$logODDs (survived)= 3.3250 + (-1.2666 * class2) + (-3.3621 * class3) + (-3.9848 * male) + (0.1617 * male * class2) + (2.3039 * male * class3)$

Women on average had a higher probability of survival over men, even
regardless of class. On average, a first class female had about a 96.5%
probability of survival on the Titanic. Second class females ended up
with a 88.7% probability of survival on the ship on average, while third
class females had about a 49.1% probability of survival. First class
males on average had highest probability of survival (34.1%) for males,
but this was still lower than females from any class. Males in second
class had about a 14.6% probability of survival, and males in third
class had a 15.2% probability of survival. Of note, the third class male
variable has a slightly higher probability of survival than males in
class 2, but that it could be due to noise as the second class male
variable is not significant. The odds (listed below) reflect a similar
pattern to the probability of survival with females having a
substantively higher odds of survival than men, regardless of class.

Overall, this logistic regression model indicates that the variables sex
and class play a role in Titanic survival. Most of these variables are
significant (p-value is below 0.05 threshold so we reject the null
hypothesis), however, there is one area where we fail to reject the null
hypothesis. The second class male variable is not significant with its
p-value of 0.7911.

### Logistic Regression

```{r}
summary(glm(survived~factor(pclass) * factor(sex), family = "binomial", data = titanic1)) 
```

### LogOdds of Survival for Class and Sex

```{r}
#Female, First Class
3.3250

#Female, Second Class
3.3250 -1.2666

#Female, Third Class 
3.3250 -3.3621

#Male, First Class 
3.3250 -3.9848 

#Male, Second Class 
3.3250 -1.2666 -3.9848 + 0.1617

#Male, Third Class 
3.3250 -3.3621 -3.9848 + 2.3039 
```

### Odds of Survival for Class and Sex

```{r}
#Odds of First Class Survival Female
exp(3.3250)

#Odds of Second Class Survival Female
exp(2.0584)

#Odds of Third Class Survival Female
exp(-0.0371)

#Odds of First Class Survival Male
exp(-0.6598)

#Odds of Second Class Survival Male
exp(-1.7647)

#Odds of Third Class Survival Male
exp(-1.718)
```

### Probability of Survival for Class and Sex

```{r}
#Probability of First Class Survival Female
exp(3.3250)/(exp(3.3250)+1) 

#Probability of Second Class Survival Female
exp(3.3250 + -1.2666)/(exp(3.3250 + -1.2666)+1) 

#Probability of Third Class Survival Female
exp(3.3250 + -3.3621)/(exp(3.3250 + -3.3621)+1) 

#Probability of First Class Survival Male
(exp(3.3250 + -3.9848)/(exp(3.3250 + -3.9848)+1))

#Probability of Second Class Survival Male
(exp(3.3250 + -1.2666 + -3.9848 + 0.1617)/(exp(3.3250 + -1.2666 + -3.9848 + 0.1617)+1))

#Probability of Third Class Survival Male
(exp(3.3250 + -3.3621 + -3.9848 + 2.3039)/(exp(3.3250 + -3.3621 + -3.9848 + 2.3039)+1))
```

### Survival Counts for Class and Sex

```{r}
#counts for class and sex
titantic_chart <- titanic1 %>% count(pclass, sex)
view(titantic_chart)

#check out survival counts for class and sex
titantic_counts <- titanic1 %>% count(survived, pclass, sex)
view(titantic_counts)
```

### Releveling of Sex to Male in Logistic Regression to Check Numbers

```{r}
summary(glm(survived~factor(pclass) * relevel(factor(sex), ref = "male"), family = "binomial", data = titanic1))

#Probability of First Class Survival Male
(exp(-0.6598)/(exp(-0.6598)+1))

#Probability of Second Class Survival Male
(exp(-0.6598 + -1.1049)/(exp(-0.6598 + -1.1049)+1))

#Probability of Third Class Survival Male
(exp(-0.6598 + -1.0582)/(exp(-0.6598 + -1.0582)+1))
```

### 
