---
title: "Homework 02"
output: html_notebook
---

# Gapminder Data - Homework 02

Import and Set Up Data

```{r}
#install.packages("gapminder") #download library

library(gapminder) #import libraries
library(tidyverse)

gapminder <- gapminder #access dataset
```

**Question \#1 (Examine Data and Set Seed)**

1\. Examine the organization of the dataset. Using the code discussed in class, randomly sample one country. Show the code used to sample. Set seed so that I could reproduce your code.

```{r}
colnames(gapminder) #see all column names
```

```{r}
set.seed(4688)
randomsample <- sample(1:1704,1)
randomsample
```

```{r}
gapminder[694,] #check which country
```

**Question \#2 (Describe Chosen Country Dynamics)**

2\.  For the country that was sampled, describe the dynamics of its population, life expectancy, and GDP per capita from the earliest to the latest year available.

```{r}
gapminder %>%
  filter(country == 'Iceland') #filter out country

#describe dynamics HERE
```

```{r}
# Plot the bar chart for Iceland
gapminder %>%
  filter(country == "Iceland") %>%
  ggplot(aes(x=year, y=pop))  + geom_line()

gapminder %>%
  filter(country == "Iceland") %>%
  ggplot(aes(x=year, y=lifeExp))  + geom_line()

gapminder %>%
  filter(country == "Iceland") %>%
  ggplot(aes(x=year, y=gdpPercap))  + geom_line()
```

```{r}
gapminder_unique<- unique(gapminder$country) #see how many unique values or countries listed 
length(gapminder_unique)
```

**Question \#3 (Examine Population, Life Expectancy, and GDP vs. World and Europe)**

3\.  How does the population, life expectancy, and GDP per capita compare with the averages for the continent and for the world as a whole in the earliest year of observation and the latest year of observation. (You will need to calculate corresponding means for this)

```{r}
#Earliest Year Averages

#World averages
gapminder_worldmin <- gapminder %>% 
  filter(year == min(year)) %>% #filter year
  summarize(avgnpop = mean(pop), avglifeexp = mean(lifeExp), avggdppercap = mean(gdpPercap))
gapminder_worldmin

#Europe averages
gapminder_europemin <- gapminder %>% 
  filter(continent == "Europe", year == min(year)) %>% #filter to Europe
  summarize(avgnpop = mean(pop), avglifeexp = mean(lifeExp), avggdppercap = mean(gdpPercap))
gapminder_europemin

#Iceland 
gapminder_icelandmin <- gapminder %>% 
  filter(country == "Iceland", year == min(year)) #filter to Iceland
gapminder_icelandmin
```

```{r}
#Latest Year Averages

#World Averages
gapminder_worldmax <- gapminder %>% 
  filter(year == max(year)) %>% #filter year
  summarize(avgnpop = mean(pop), avglifeexp = mean(lifeExp), avggdppercap = mean(gdpPercap))
gapminder_worldmax

#Europe Averages
gapminder_europemax <- gapminder %>% 
  filter(continent == "Europe", year == max(year)) %>% #filter to Europe and year
  summarize(avgnpop = mean(pop), avglifeexp = mean(lifeExp), avggdppercap = mean(gdpPercap))
gapminder_europemax

#Iceland 
gapminder_icelandmax <- gapminder %>% 
  filter(country == "Iceland", year == max(year)) 
gapminder_icelandmax
```

**Question \#4 (Standard Deviation and Z-Score)**

4\. Calculate the standard deviation for the continent and for the world for each indicator for the first and last year of observations. How many standard deviations away from the mean was your country in the first year of observation? In the last year of observation? What does it say about the development of your country?

```{r}
#Earliest Year Standard Deviation

#World SD
gapminder_worldmin_sd <- gapminder %>% 
  filter(year == min(year)) %>% #filter year
  summarize(sdpop = sd(pop), sdexp = sd(lifeExp), sdgdpercap = sd(gdpPercap))
gapminder_worldmin_sd

#Europe SD
gapminder_europemin_sd <- gapminder %>% 
  filter(continent == "Europe", year == min(year)) %>% #filter to Europe and year
  summarize(sdpop = sd(pop), sdexp = sd(lifeExp), sdgdpercap = sd(gdpPercap))
gapminder_europemin_sd
```

```{r}
#Latest Year Standard Deviation

#World SD
gapminder_worldmax_sd <- gapminder %>% 
  filter(year == max(year)) %>% #filter year
  summarize(sdpop = sd(pop), sdexp = sd(lifeExp), sdgdpercap = sd(gdpPercap))
gapminder_worldmax_sd

#Europe SD
gapminder_europemax_sd <- gapminder %>% 
  filter(continent == "Europe", year == max(year)) %>% #filter to Europe and year
  summarize(sdpop = sd(pop), sdexp = sd(lifeExp), sdgdpercap = sd(gdpPercap))
gapminder_europemax_sd
```

```{r}
#Z-score for Iceland vs. World and Europe - Population


#Iceland vs. World z-score
zscore_vs_world <- gapminder %>% 
  group_by(year) %>% 
  mutate(avgpop = mean(pop), sdpop = sd(pop)) %>% 
  filter(country == "Iceland") %>% 
  mutate(zscore = ((pop-avgpop)/sdpop))
zscore_vs_world

zscore_vs_world %>%
  filter(year == "1952") #ASK ABOUT FILTERING MIN and MAX YEAR

zscore_vs_world %>%
  filter(year == "2007") 

#Iceland vs. Europe z-score
zscore_vs_europe <- gapminder %>% 
  group_by(year) %>% 
  filter(continent == "Europe") %>%
  mutate(avgpop = mean(pop), sdpop = sd(pop)) %>% 
  filter(country == "Iceland") %>% 
  mutate(zscore = ((pop-avgpop)/sdpop))
zscore_vs_europe

zscore_vs_europe %>%
  filter(year == "1952") #ASK ABOUT FILTERING MIN and MAX YEAR

zscore_vs_europe %>%
  filter(year == "2007")

```

```{r}
#Z-score for Iceland vs. World and Europe - Life Expectancy

#Iceland vs. World z-score
zscore_vs_world <- gapminder %>% 
  group_by(year) %>% 
  mutate(avglifeex = mean(lifeExp), sdlifeex = sd(lifeExp)) %>% 
  filter(country == "Iceland") %>% 
  mutate(zscore = ((lifeExp-avglifeex)/sdlifeex))
zscore_vs_world

zscore_vs_world %>%
  filter(year == "1952") #ASK ABOUT FILTERING MIN and MAX YEAR

zscore_vs_world %>%
  filter(year == "2007") 

#Iceland vs. Europe z-score
zscore_vs_europe <- gapminder %>% 
  group_by(year) %>% 
  filter(continent == "Europe") %>%
  mutate(avglifeex = mean(lifeExp), sdlifeex = sd(lifeExp)) %>% 
  filter(country == "Iceland") %>% 
  mutate(zscore = ((lifeExp-avglifeex)/sdlifeex))
zscore_vs_europe

zscore_vs_europe %>%
  filter(year == "1952") #ASK ABOUT FILTERING MIN and MAX YEAR

zscore_vs_europe %>%
  filter(year == "2007")

```

```{r}
#Z-score for Iceland vs. World and Europe - GDP Per Capita

#Iceland vs. World z-score
zscore_vs_world <- gapminder %>% 
  group_by(year) %>% 
  mutate(avggdppercap = mean(gdpPercap), sdgdppercap = sd(gdpPercap)) %>% 
  filter(country == "Iceland") %>% 
  mutate(zscore = ((gdpPercap-avggdppercap)/sdgdppercap))
zscore_vs_world

zscore_vs_world %>%
  filter(year == "1952") #ASK ABOUT FILTERING MIN and MAX YEAR

zscore_vs_world %>%
  filter(year == "2007") 

#Iceland vs. Europe z-score
zscore_vs_europe <- gapminder %>% 
  group_by(year) %>% 
  filter(continent == "Europe") %>%
  mutate(avggdppercap = mean(gdpPercap), sdgdppercap = sd(gdpPercap)) %>% 
  filter(country == "Iceland") %>% 
  mutate(zscore = ((gdpPercap-avggdppercap)/sdgdppercap))
zscore_vs_europe

zscore_vs_europe %>%
  filter(year == "1952") #ASK ABOUT FILTERING MIN and MAX YEAR

zscore_vs_europe %>%
  filter(year == "2007")
```

**Question \#5 (How Iceland Compares in Latest Year)**

5\. In the last year of observations, how many countries outperformed your country in terms of each indicator? What was the best and the worst-performing country in the world and in the continent of your country?

```{r}
gapminder_lastyear <- gapminder %>% 
  filter(year == max(year)) #filter year

gapminder_iceland <- gapminder %>% 
  filter(country == "Iceland", year == max(year)) 

table(gapminder_iceland$pop>gapminder_lastyear$pop)
table(gapminder_iceland$lifeExp>gapminder_lastyear$lifeExp)
table(gapminder_iceland$gdpPercap>gapminder_lastyear$gdpPercap)


#FIND RANK?
```

```{r}
#Best and Worst Population
gapminder %>%
  filter(year == max(year)) %>%
  filter(pop == min(pop))

gapminder %>%
  filter(year == max(year)) %>%
  filter(pop == max(pop))

gapminder %>%
  filter(continent == "Europe", year == max(year)) %>%
  filter(pop == min(pop))

gapminder %>%
  filter(continent == "Europe", year == max(year)) %>%
  filter(pop == max(pop))

```

```{r}
#Best and Worst Life Expectancy
gapminder %>%
  filter(year == max(year)) %>%
  filter(lifeExp == min(lifeExp))

gapminder %>%
  filter(year == max(year)) %>%
  filter(lifeExp == max(lifeExp))

gapminder %>%
  filter(continent == "Europe", year == max(year)) %>%
  filter(lifeExp == min(lifeExp))

gapminder %>%
  filter(continent == "Europe", year == max(year)) %>%
  filter(lifeExp == max(lifeExp))


```

```{r}
#Best and Worst GDP Per Captia
gapminder %>%
  filter(year == max(year)) %>%
  filter(gdpPercap == min(gdpPercap))

gapminder %>%
  filter(year == max(year)) %>%
  filter(gdpPercap == max(gdpPercap))

gapminder %>%
  filter(continent == "Europe", year == max(year)) %>%
  filter(gdpPercap == min(gdpPercap))

gapminder %>%
  filter(continent == "Europe", year == max(year)) %>%
  filter(gdpPercap == max(gdpPercap))
```

**Question \#6 (Countries Improvement Over Time)**

6\. In how many countries was the improvement over time better than in your country in the world? In your continent?

```{r}
iceland_max <- gapminder %>%
  filter(country == "Iceland", year == max(year))

iceland_min <- gapminder %>%
  filter(country == "Iceland", year == min(year))

iceland_pop_change <-iceland_max$pop - iceland_min$pop 
iceland_lifeExp_change <- iceland_max$lifeExp - iceland_min$lifeExp 
iceland_gdppercap_change <- iceland_max$gdpPercap - iceland_min$gdpPercap 


gapminder_2007 <- gapminder %>%
  filter(year== max(year))
gapminder_2007

gapminder_1952 <- gapminder %>%
  filter(year== min(year))
gapminder_1952

table(iceland_pop_change> (gapminder_2007$pop-gapminder_1952$pop))
table(iceland_pop_change> (gapminder_2007$pop-gapminder_1952$pop))
table(iceland_pop_change> (gapminder_2007$pop-gapminder_1952$pop))

 

```

**Question \#7 (Total Population Each Year)**

7\.  What was the total population of the continent each year?  What was the total population of the world?

```{r}
#Total Population of Europe Each Year
gapminder_pop_europe <- gapminder %>%
  filter(continent == "Europe") %>%
  group_by(year)%>%
  summarize(by_year = sum(pop))
gapminder_pop_europe
```

```{r}
#Total Population of World
gapminder_pop_world <- gapminder %>%
  group_by(year)%>%
  summarize(by_year = sum(pop))
gapminder_pop_world
```

**Question \#8 (Total GDP World)**

8\. What was the total GDP (not per capita) of the world in each year according to the data?

```{r}
#GPD/Population = GDP Per Capita 
gapminder_GDP <- gapminder %>%
  mutate(gdp = pop*gdpPercap)%>%
  group_by(year) %>% 
  summarize(by_year = sum(pop))
gapminder_GDP
```

**Question \#9**

9\.  Write a short paragraph describing your country's development in history compared to other countries in its continent and the world as a whole. Make sure to write it in a way so that someone who doesn't know statistics could understand it. 

Iceland based o \~\~ finished answer here
